A Sound and Complete Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2024-01-30 paper/COLA/paper.tex
Some fixes and missing pieces

\author{Paulo Torrens}
\ead{paulotorrens@gnu.org}
\address{Santa Catarina State University}

@@

%
\begin{abstract}
  This work describes an encoding of modal logics using the Coq proof assistant.
  Our formalization differs from previous attempts by providing a uniform
  representation of several systems for modal logic using Coq's rich type
  structure. We illustrate the usefulness of our library in a formalization of
  L\"ob's theorem which closely follows a classical proof of this result.
\end{abstract}

@@

The start of modal logics can be traced back to Lewis and his definition of 
logical systems with strict implication instead of material implication \cite{Lewis1918}.
Based in this work, independently of classical propositional logic,
defined the systems $\mathsf{S1}$--$\mathsf{S5}$~\cite{Lewis1932}, where $\mathsf{S4}$ and $\mathsf{S5}$
are well known normal modal systems. It was, in fact, G{\"o}del that had the idea 
to build a modal logic system as an extension of the classical propositional calculus, 
when constructing a system similar to $\mathsf{S4}$ in \cite{Godel1933}.

Despite that the original motivation of modal systems was to deal with strict implication, 
for many years modal logic was seen as ``the logic of necessity and possibility''. 
Only after the introduction of Kripke semantics they were also seen as ``simple yet
expressive languages for talking about relational structures''~\cite{Blackburn-ModalLogics}. 
A great multitude of modal systems, with various kinds of modal operators, each of them with 
different meanings or properties, have then been constructed in order to provide effective
formalisms for talking about time, space, knowledge, beliefs, actions, 
obligations, among others.

The interpretation of modal logic simply as ``the logic of necessity'' restricts 
not only its applications but also the range of modal systems. In fact, modal logic 
can also be seen as ``the study of \emph{modalities} -- logical operations that qualify 
assertions about the truth of statements''~\cite{Goldblatt-MathofModality}. In this sense, 
the necessitation rule (from $\varphi$ infer $\Box\varphi$) may be relaxed. This brings 
a number of modal systems known as non-normal, whose semantics were defined only in the 
1970's by Montague~\cite{Montague1970} and Scott~\cite{Scott_AdviceModal1970}.

Such wide range of highly expressive logical languages is the ideal tool to 
formalize the many complexities of modern advanced software systems, 
such as those found in cars, airplanes, medical devices and banking applications.
In combining said modal logics with interactive proof assistants,
such as Coq~\cite{manual_coq} and Isabelle/HOL~\cite{manual_isabelle}, both of
which have been applied with success in non-trivial projects like a certified compiler 
for C programming language~\cite{Leroy09} and the kernel of seL4 operating
system~\cite{Klein10}, it is possible to formalize and verify highly complex computer
systems.

Although there are some dedicated provers for modal logics~\cite{Gleissner17,Mora11},
we believe that using mature tools, like a proof assistant, is a better option for
verifying facts using modal logics. We justify this choice by: 1) the recent
successful histories on verification of software and
mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) there is a growing
community of developers using such tools thanks to 
textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
newcomers to join the game of formal verification; and 3) the custom automation
facilities supported by current proof assistants allows for trivial facts to
be proven using decision procedures, relieving the user from writing proof
scripts for formulas that can be checked automatically.

@@

\begin{itemize}
  \item We define a \emph{deep embedding} of modal logic syntax in Coq and
    define its Kripke semantics.
  \item We model the Hilbert deductive system as a Coq inductive type that is
    parameterized by different axiom systems like K, D, B, T, K4, K5, S4, S5 and GL.
  \item We formally verify several frame correspondence results.
  \item We prove soundness and completeness for the Hilbert deductive system. For the
        completeness proof we follow the so-called Henkin-style and it closely follows
        the proofs presented in classical modal logics textbooks~\cite{Blackburn-ModalLogics}.
  \item As an example of usage, we show a proof of L\"ob's theorem using the deductive system
    developed in our library.
\end{itemize}

This article is an extended version of our SBLP 2022 work~\cite{Silveira22}. In this paper we include
a detailed discussion about our formalization of a completeness proof for the Hilbert deductive
system following the strategies found in traditional modal logic books~\cite{Blackburn-ModalLogics}.
The rest of this work is organized as follows: Section~\ref{sec:background} provides
brief introduction on modal logics and the Coq proof assistant. Details about
the implementation of our Coq library are presented in
Section~\ref{sec:implementation}. A case study in proving L\"ob's theorem is
the subject of Section~\ref{sec:case}. Related works are discussed
on Section~\ref{sec:related} and Section~\ref{sec:conclusion} draws some conclusions and
provide some pointers for future work.

All source code in this article has been formalized in Coq version 8.15 and
is available at \url{https://github.com/funcao/ModalLibrary/} under a permissive free software license.

@@

\subsection{An overview of Coq proof assistant.}

Coq is a proof assistant based on the calculus of inductive
constructions (CIC)~\cite{manual_cic}, a higher-order typed
$\lambda$-calculus extended with inductive definitions.  Theorem
proving in Coq follows the ideas of the so-called
``BHK-in\-ter\-pre\-ta\-tion''\footnote{Abbreviation of Brouwer, Heyting,
  Kolmogorov, de Bruijn and Martin-L\"of Correspondence. This is also
  known as the Curry-Howard isomorphism.}, where types represent
logical formulas, $\lambda$-terms represent proofs
\cite{Bertot10} and the task of checking if a piece of text is a
proof of a given formula corresponds to checking if the term that
represents the proof has the type corresponding to the given formula.

In order to make writing proofs easier, Coq provides
\emph{tactics}, which are commands that can be used to construct proof
terms in a more user-friendly way. As an example, consider the task of proving the following 
formula of the propositional logic:
\[
(A \to B)\to (B\to C) \to A \to C
\]
In Coq, such theorem can be expressed as:
\begin{minted}{coq}
Section EXAMPLE.
  Variables A B C: Prop.
  Theorem example: (A -> B) -> (B -> C) -> A -> C.
  Proof.
    intros H H' HA.
    apply H'.
    apply H.
    assumption. 
  Qed.
End EXAMPLE.
\end{minted}
In the above, we have defined a Coq section named
\mintinline{coq}{EXAMPLE}\footnote{In Coq, we can use sections to delimit the
  scope of local variables.} which declares variables \mintinline{coq}{A},
\mintinline{coq}{B} and \mintinline{coq}{C} as being propositions (i.e. with type
\mintinline{coq}{Prop}). Tactic \mintinline{coq}{intros} introduces variables
\mintinline{coq}{H}, \mintinline{coq}{H'} and \mintinline{coq}{HA} into the typing context,
respectively with types \mintinline{coq}{A -> B}, \mintinline{coq}{B -> C}, and
\mintinline{coq}{A}, and leaves goal \mintinline{coq}{C} to be proved. Tactic
\mintinline{coq}{apply}, used with a term \mintinline{coq}{t}, generates goal
\mintinline{coq}{P}
when there exists \mintinline{coq}{t: P -> Q} in the typing context and the
current goal is \mintinline{coq}{Q}. Thus, \mintinline{coq}{apply H'} changes the goal
from \mintinline{coq}{C} to \mintinline{coq}{B} and \mintinline{coq}{apply H}
changes the goal to \mintinline{coq}{A}. Tactic \mintinline{coq}{assumption}
traverses the typing context to find a hypothesis that matches the goal,
concluding the proof.

@@

\subsection{An overview of modal logic.} Modal logics are designed for reasoning
about truth across various --- abstract --- worlds. In such logics, a
proposition may be true in some world, but false in another one. The systems of
modal logic considered in this work extend traditional propositional logic
with two operators on propositions: $\Box$ (box) and $\Diamond$ (diamond).
The syntax for modal logic formulas is defined by the following context free
grammar (where
$\mathcal{P} = \{A,B,C,...\}$ is the countable set of propositional variables):
\[
  \begin{array}{lcl}
    \varphi & ::= & \mathcal{P}\,|\,\neg\varphi\,|\,\varphi\land\varphi\,|\,\varphi\lor\varphi\,|\,\varphi\to\varphi\,|\,\Box\varphi\,|\,\Diamond\varphi
  \end{array}  
\]
Intuitively, the semantics of the operator $\Box$ is similar to an universal
quantifier, since it expresses \emph{necessity}: $\Box\,\varphi$ is true in the
current world iff $\varphi$ is true in every world accessible from the current
one. Similarly, the semantics of operator
$\Diamond$ is that a formula $\Diamond\varphi$ is true in the current
world iff $\varphi$ is \emph{possible}, i.e. true in at least some world
accessible from the current world. In the remaining of the paper, we also write
$\phi_1\leftrightarrow\phi_2$ to mean $\phi_1\to\phi_2\land\phi_2\to\phi_1$.

Formally, we can only interpret a formula with respect to a \emph{model}, which is
defined as follows.
A \emph{frame} $\mathcal{F} = \langle W, R \rangle$ consists of a pair formed by
a non-empty set of worlds, $W$, and a binary accessibility relation between
worlds, $R \subseteq W \times W$. A pair $w_1Rw_2$ denotes that world $w_2$ is
accessible from $w_1$. A model $\mathcal{M} = \langle \mathcal{F}, \mathcal{V} \rangle$
is formed by a frame $\mathcal{F}$ and a total labeling function
$\mathcal{V} : \mathcal{P}\to 2^W$, which indicates the set of worlds where each propositional variable holds, i.e. when $w \in \mathcal{V}(A)$
for some $A \in\mathcal{P}$, we say that $A$ holds in $w$. The satisfiability of a formula
in a given world is defined by inductive relation $\mathcal{M}\,;\,w\models \varphi$ as
follows (notation $\mathcal{M}\,;\,w\not\models \varphi$ denotes
that $\mathcal{M}\,;\,w\models \varphi$ does not hold):

\begin{figure}[H]
\begin{enumerate}
  % \item $\mathcal{M}\,;\,w\not\models \bot$ and $\mathcal{M}\,;\,w\models \top$;
  \item $\mathcal{M}\,;\,w\models A$ iff $w \in \mathcal{V}(A)$;
  \item $\mathcal{M}\,;\,w\models \neg\varphi$ iff $\mathcal{M}\,;\,w\not\models\varphi$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \land \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ and $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \lor \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \to \varphi_2$ iff
    $\mathcal{M}\,;\,w\not\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models\Box\varphi$ iff $\forall y \in W. (w R y \to \mathcal{M}\,;\,y \models\varphi)$;
  \item $\mathcal{M}\,;\,w\models\Diamond\varphi$ iff $\exists y \in W. (w R y \land \mathcal{M}\,;\,y \models\varphi)$;
  \end{enumerate}
  \caption{Kripke semantics}
  \label{fig:kripke}
\end{figure}
We say that a formula $\varphi$ is satisfiable in a model $\mathcal{M}$,
denoted by $\mathcal{M}\models \varphi$, if $\forall w \in W. (M\,;w\,\models \varphi)$ holds.
Similarly, $\varphi$ is valid in a frame $\mathcal{F}$, written $\mathcal{F}\models\varphi$, if
it holds on all models $\mathcal{M} = \langle \mathcal{F},\mathcal{V}\rangle$, i.e.
for any $\mathcal{M}$, we have that $\mathcal{M}\models\varphi$. A formula
is valid, written $\models \varphi$, if it is valid in any frame $\mathcal{F}$.

Using the semantics, we can define the concept of \emph{entailment}: given a set
of formulas $\Gamma$ and a formula $\varphi$, we say that $\Gamma$ entails
$\varphi$, written $\Gamma\models\varphi$, iff $\forall \mathcal{M} . ((\forall \varphi' \in\Gamma .
\mathcal{M} \models \varphi') \to \mathcal{M} \models \varphi)$.

In this work, we consider Hilbert deductive systems for modal logic.
Essentially, the considered system adds axioms for the modalities
to the set of axioms for classical propositional logic.
As examples of modalities axioms we mention K, which express that $\Box$ distributes
over implication, and possibility which states that $\Diamond$ distributes
over disjunction.
\[
  \begin{array}{ll}
    K & \Box(\varphi_1 \to \varphi_2) \to (\Box \varphi_1 \to \Box \varphi_2)\\
    Possibility & \Diamond (\varphi_1 \lor \varphi_2) \to
                           (\Diamond \varphi_1 \lor \Diamond \varphi_2)\\
  \end{array}
\]
Apart from the axioms, we also need \emph{modus ponens} and necessitation rules.

@@

A central aspect in the study of modal logics is the so-called \emph{frame
  correspondence}, which is a relationship between the
accessibility relation and the truth of certain formulas in frames containing
it. As an example, consider a frame $\mathcal{F} = \langle W,R
\rangle$ in which the relation $R$ is reflexive: $\forall w \in W. (wRw)$, i.e.
every world is accessible from itself. In such frame when $\Box A$ holds in
a world $w$, $w$ itself is among the accessible worlds in which $A$ must be true.
In this way, if the accessibility relation $R$ is reflexive, then whatever world
$w$ and formula $\varphi$ we take, the formula $\Box \varphi \to \varphi$ will
be true there. Examples of properties that characterize certain formula schemas are
presented in the table below.

@@

In this section, we describe the implementation details of our modal logic
library in Coq. We start by describing the structure of formula syntax and the
representation of frames and models, needed to define the Kripke semantics.
Next, we define a modular Hilbert-style deductive system. The modularization is
obtained by parameterizing it by a set of axioms used. This allows us to
generalize results about derivations to several deductive systems, independently
of the axioms used. We conclude by discussing the soundness and the frame
correspondence proofs.

@@

To implement our library, we use a \emph{deep embedding} of
formulas, i.e. we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive formula: Set :=
| Lit    : nat -> formula
| Neg    : formula -> formula
| Box    : formula -> formula
| Dia    : formula -> formula
| And    : formula -> formula -> formula
| Or     : formula -> formula -> formula
| Implies: formula -> formula -> formula.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logic. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$. Furthermore, the type \mintinline{coq}{theory} is defined as a function
between \mintinline{coq}{formula} and Coq's \mintinline{coq}{Prop}s.

@@

Before defining the Kripke semantics, we need to represent the model used to
interpret a modal logic formula. We encode a frame by a record formed by
a type \mintinline{coq}{W: Set} to represent the set of worlds and an
accessibility relation denoted by a type \mintinline{coq}{R: W -> W -> Prop}.
\begin{minted}{coq}
Record Frame: Type := {
  W: Set;
  R: W -> W -> Prop;
}.
\end{minted}
A model is then denoted by another record, formed by a frame and a labeling function,
which assigns the truth value of a variable at a given world.
\begin{minted}{coq}
Record Model: Type := {
  F: Frame; 
  v: nat -> (W F) -> Prop; 
}.
\end{minted}
Note that the type of the labeling function \mintinline{coq}{v} ensures that it
refers to a world from the set of possible worlds in the frame
\mintinline{coq}{F}\footnote{We recall that \mintinline{coq}{W F} is Coq's syntax for
accessing field \mintinline{coq}{W} from the record \mintinline{coq}{F}.}. The encoding of the Kripke semantics is done by a Coq function
that implements the relation of Figure~\ref{fig:kripke}. We omit the code
for propositional logics connectives for brevity. The interesting parts are for
variables (constructor \mintinline{coq}{Lit}) which uses the labeling function
\mintinline{coq}{v} from the model \mintinline{coq}{M}. The meaning of
modalities $\Box$ and $\Diamond$ is represented by quantifying over worlds
using the accessibility relation $R$. 
\begin{minted}{coq}
Fixpoint formula_valuation (M: Model) (w: W (F M))
                           (phi: formula): Prop :=
  match phi with
  | Lit     x
     => v M x w (* should be inhabited, i.e., true *)
  | Box     psi
     => forall w': W (F M), R (F M) w w' ->
                            formula_valuation M w' psi
  | Dia     psi
     => exists w': W (F M), R (F M) w w' /\
                            formula_valuation M w' psi
  (* some code omitted *)
  end.
\end{minted}
In order to ease the task of writing formulas and its semantics properties, we
define Coq notations that mimic the standard way of writing them.

@@

A key property of the Hilbert deductive system as formalized
in our library is its soundness (through system K). If a formula $\varphi$ is derived from a
context $\Gamma$, i.e. $\Gamma\vdash\varphi$, then $\varphi$ is entailed within the deep embedding
from the set of hypotheses $\Gamma$, i.e. $\Gamma\models\varphi$. The Coq statement for the soundness
result is given as follow:

\begin{minted}{coq}
Theorem soundness : forall G p, (K; G |-- p) -> (G ||= p).
\end{minted}

>>>>

\author{Paulo Torrens}
\ead{paulotorrens@gnu.org}
\address{University of Kent}

@@

%
\begin{abstract}
  This work describes a deep-embedding formalization of modal logic systems using the Coq proof assistant.
  Our formalization differs from previous attempts by providing a uniform
  representation of several systems for modal logic using Coq's rich type
  structure by quantifying over the set of axioms, for which we prove soundness and completeness. Finally, we motivate the use of our library in a formalization of
  L\"ob's theorem which closely follows a classical proof of this result.
\end{abstract}

@@

The origins of modal logic can be traced back to Lewis and his definition of
logical systems with strict implication instead of material implication \cite{Lewis1918}.
Based on this work, and independently of classical propositional logic, he
defined the systems $\mathsf{S1}$--$\mathsf{S5}$~\cite{Lewis1932}, where $\mathsf{S4}$ and $\mathsf{S5}$
are now well known normal modal systems. It was, in fact, G{\"o}del that had the idea
to build a modal logic system as an extension of the classical propositional calculus,
when constructing a system similar to $\mathsf{S4}$ in \cite{Godel1933}.

Despite that the original motivation of modal systems was to deal with strict implication,
for many years modal logic has been seen as ``the logic of necessity and possibility''.
Only after the introduction of Kripke semantics they were also seen as ``simple yet
expressive languages for talking about relational structures''~\cite{Blackburn-ModalLogics}.
A great multitude of modal systems, with various kinds of modal operators, each of them with
different meanings or properties, have then been constructed in order to provide effective
formalisms for talking about time, space, knowledge, beliefs, actions,
obligations, among others, by axiomatizing different structures.

We note that the interpretation of modal logic simply as ``the logic of necessity'' restricts
not only its applications but also the possible range of modal systems. In fact, modal logic
can also be seen as ``the study of \emph{modalities} -- logical operations that qualify
assertions about the truth of statements''~\cite{Goldblatt-MathofModality}. In this sense,
the standard necessitation rule (i.e., that from $\varphi$ we may infer $\Box\varphi$) may be actually relaxed. This brings
a number of modal systems known as non-normal, whose semantics were defined only in the
1970's by Montague~\cite{Montague1970} and Scott~\cite{Scott_AdviceModal1970}.

Such wide range of highly expressive logical languages is the ideal tool to
formalize the many non-trivial properties of modern advanced software systems,
such as those found in cars, airplanes, medical devices and banking applications.
By combining said modal logics with interactive proof assistants,
such as Coq~\cite{manual_coq} and Isabelle/HOL~\cite{manual_isabelle}, both of
which have been applied with success in non-trivial projects like a certified compiler
for C programming language~\cite{Leroy09} and the kernel of an operating
system~\cite{Klein10},
% it is possible to formalize and verify highly complex computer systems.
we argue we may achieve betters ways of formally describing and verifying complex programs.

Although there are some dedicated provers for modal logics~\cite{Gleissner17,Mora11},
we believe that using mature tools, like a proof assistant, would be a better option for
verifying facts using modal logics. We justify this choice by: 1) the recent
successful histories on verification of software and
mathematics~\cite{Leroy09,Klein10,Gonthier13,Gonthier08}; 2) the growing
community of developers using such tools thanks to
textbooks on the subject~\cite{Nipkow14,Bertot10,Chlipala13} which allows
newcomers to join the game of formal verification; and 3) the custom automation
facilities supported by current proof assistants allows for trivial facts to
be proven using decision procedures, relieving the user from writing proof
scripts for formulas that can be checked automatically.

@@

\begin{itemize}
  \item We define a \emph{deep embedding} of modal logic syntax in Coq and
    define its Kripke semantics.
  \item We model the Hilbert-style deductive system as a Coq inductive type that is
    parameterized by the set of axioms, allowing to derive systems like K, D, B, T, K4, K5, S4, S5 and GL.
  \item We formally verify several frame correspondence results.
  \item We prove both soundness and completeness for the deductive system with regard to its embedding within Coq's own logic (the calculus of inductive constructions).

  % For the
  %       completeness proof we follow the so-called Henkin-style and it closely follows
  %      the proofs presented in classical modal logics textbooks~\cite{Blackburn-ModalLogics}.
  \item As an example of usage, and in order to motivate the use of our library, we show a proof of L\"ob's theorem using the deductive system
    developed.
\end{itemize}

This article is an extended version of our SBLP 2022 work~\cite{Silveira22}. In this paper we also include
a detailed discussion about our formalization of a completeness proof for the Hilbert-style deductive
system following the strategies found in traditional modal logic books~\cite{Blackburn-ModalLogics}.
The rest of this work is organized as follows: Section~\ref{sec:background} provides
brief introduction on modal logics and the Coq proof assistant. Details about
the implementation of our Coq library are presented in
Section~\ref{sec:implementation}. A case study in proving L\"ob's theorem is
the subject of Section~\ref{sec:case}. Related works are discussed
on Section~\ref{sec:related} and Section~\ref{sec:conclusion} draws some conclusions and
provide some pointers for future work.
%
All source code in this article has been formalized in Coq version 8.15 and
is available at \url{https://github.com/funcao/LML/} under a permissive free software license.

@@

\subsection{A brief overview of Coq proof assistant.}

Coq is a proof assistant based on the calculus of inductive
constructions ~\cite{manual_cic}, a higher-order dependently typed
$\lambda$-calculus extended with inductive definitions.  Theorem
proving in Coq follows the ideas of the so-called
``BHK-in\-ter\-pre\-ta\-tion''\footnote{Abbreviation of Brouwer, Heyting,
  Kolmogorov, de Bruijn and Martin-L\"of Correspondence. This is also
  known as the Curry-Howard isomorphism.}, where types represent
logical formulas, $\lambda$-terms represent proofs
\cite{Bertot10} and the task of checking if a piece of text is a
proof of a given formula corresponds to checking if the term that
represents the proof has the type corresponding to the given formula.

In order to make writing proofs easier, Coq provides
\emph{tactics}, which are commands that can be used to construct proof
terms in a more user-friendly way. As an example, consider the task of proving the following
formula of the propositional logic:
\[
(A \to B)\to (B\to C) \to A \to C
\]
In Coq, such theorem can be expressed as:
\begin{minted}{coq}
Section EXAMPLE.
  Variables A B C: Prop.
  Theorem example: (A -> B) -> (B -> C) -> A -> C.
  Proof.
    intros H H' HA.
    apply H'.
    apply H.
    assumption.
  Qed.
End EXAMPLE.
\end{minted}
In the above, we have defined a Coq section named
\mintinline{coq}{EXAMPLE}\footnote{In Coq, we can use sections to delimit the
  scope of local variables.} which declares variables \mintinline{coq}{A},
\mintinline{coq}{B} and \mintinline{coq}{C} as being propositions (i.e., with type
\mintinline{coq}{Prop}). Tactic \mintinline{coq}{intros} introduces variables
\mintinline{coq}{H}, \mintinline{coq}{H'} and \mintinline{coq}{HA} into the typing context,
respectively with types \mintinline{coq}{A -> B}, \mintinline{coq}{B -> C}, and
\mintinline{coq}{A}, and leaves goal \mintinline{coq}{C} to be proved. Tactic
\mintinline{coq}{apply}, used with a term \mintinline{coq}{t}, generates goal
\mintinline{coq}{P}
when there exists \mintinline{coq}{t: P -> Q} in the typing context and the
current goal is \mintinline{coq}{Q}. Thus, \mintinline{coq}{apply H'} changes the goal
from \mintinline{coq}{C} to \mintinline{coq}{B} and \mintinline{coq}{apply H}
changes the goal to \mintinline{coq}{A}. Tactic \mintinline{coq}{assumption}
traverses the typing context to find a hypothesis that matches the goal,
concluding the proof.

@@

\subsection{An overview of modal logic.} Modal logics are designed for reasoning
about truth across various --- abstract --- worlds. In such logics, a
proposition may be true in some world, but false in another one. The systems of
modal logic considered in this work extend traditional propositional logic
with two operators on propositions: $\Box$ (box) and $\Diamond$ (diamond).
The syntax for modal logic formulas is defined by the following context free
grammar (where
$\mathcal{P} = \{A,B,C,...\}$ is the countable set of propositional variables):
\[
  \begin{array}{lcl}
    \varphi & ::= & \mathcal{P}\,|\,\neg\varphi\,|\,\varphi\land\varphi\,|\,\varphi\lor\varphi\,|\,\varphi\to\varphi\,|\,\Box\varphi\,|\,\Diamond\varphi
  \end{array}
\]
Intuitively, the semantics of the operator $\Box$ is similar to a universal
quantifier, since it expresses \emph{necessity}: $\Box\,\varphi$ is true in the
current world iff $\varphi$ is true in every world accessible from the current
one. Similarly, the semantics of operator
$\Diamond$ is that a formula $\Diamond\varphi$ is true in the current
world iff $\varphi$ is \emph{possible}, i.e., true in at least some world
accessible from the current world. In the remaining of the paper, we also write
$\phi_1\leftrightarrow\phi_2$ to mean $\phi_1\to\phi_2\land\phi_2\to\phi_1$.

Formally, we can only interpret a formula with respect to a \emph{model}, which is
defined as follows.
A \emph{frame} $\mathcal{F} = \langle W, R \rangle$ \linebreak consists of a pair formed by
a non-empty set of worlds, $W$, and a binary accessibility relation between
worlds,
\linebreak
$R \subseteq W \times W$. A pair $w_1Rw_2$ denotes that world $w_2$ is
accessible from $w_1$. A model $\mathcal{M} = \langle \mathcal{F}, \mathcal{V} \rangle$
is formed by a frame $\mathcal{F}$ and a total labeling function
$\mathcal{V} : \mathcal{P}\to 2^W$, which indicates the set of worlds where each propositional variable holds, i.e., when $w \in \mathcal{V}(A)$
for some $A \in\mathcal{P}$, we say that $A$ holds in $w$. The satisfiability of a formula
in a given world is defined by inductive relation $\mathcal{M}\,;\,w\models \varphi$ as
follows (notation $\mathcal{M}\,;\,w\not\models \varphi$ denotes
that $\mathcal{M}\,;\,w\models \varphi$ does not hold):

\begin{figure}[H]
\begin{enumerate}
  % \item $\mathcal{M}\,;\,w\not\models \bot$ and $\mathcal{M}\,;\,w\models \top$;
  \item $\mathcal{M}\,;\,w\models A$ iff $w \in \mathcal{V}(A)$;
  \item $\mathcal{M}\,;\,w\models \neg\varphi$ iff $\mathcal{M}\,;\,w\not\models\varphi$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \land \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ and $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \lor \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \to \varphi_2$ iff
    $\mathcal{M}\,;\,w\not\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models\Box\varphi$ iff $\forall y \in W. (w R y \to \mathcal{M}\,;\,y \models\varphi)$;
  \item $\mathcal{M}\,;\,w\models\Diamond\varphi$ iff $\exists y \in W. (w R y \land \mathcal{M}\,;\,y \models\varphi)$.
  \end{enumerate}
  \caption{Kripke semantics}
  \label{fig:kripke}
\end{figure}
We say that a formula $\varphi$ is satisfiable in a model $\mathcal{M}$,
denoted by $\mathcal{M}\models \varphi$, if $\forall w \in W. (M\,;w\,\models \varphi)$ holds.
Similarly, $\varphi$ is valid in a frame $\mathcal{F}$, written $\mathcal{F}\models\varphi$, if
it holds on all models $\mathcal{M} = \langle \mathcal{F},\mathcal{V}\rangle$, i.e.,
for any $\mathcal{M}$, we have that $\mathcal{M}\models\varphi$. A formula
is valid, written $\models \varphi$, if it is valid in any frame $\mathcal{F}$. Using the above semantics, we can define the concept of \emph{entailment}: given a set
of formulas $\Gamma$ and a formula $\varphi$, we say that $\Gamma$ entails
$\varphi$, written $\Gamma\models\varphi$, iff $\forall \mathcal{M} . ((\forall \varphi' \in\Gamma .
\mathcal{M} \models \varphi') \to \mathcal{M} \models \varphi)$.
In this definition, if $\Gamma$ entails $\varphi$ it means that $\varphi$ is a global semantic consequence of $\Gamma$, which will be further elaborated on Section~\ref{sec:completeness}.

In this work, we consider Hilbert deductive systems for modal logic.
Essentially, the considered system add axioms for the modalities
to the axioms for classical propositional logic.
As examples of modalities axioms we mention $K$, which express that $\Box$ distributes
over implication, and possibility, which states that $\Diamond$ distributes
over disjunction.
\[
  \begin{array}{ll}
    K & \Box(\varphi_1 \to \varphi_2) \to (\Box \varphi_1 \to \Box \varphi_2)\\
    Possibility & \Diamond (\varphi_1 \lor \varphi_2) \to
                           (\Diamond \varphi_1 \lor \Diamond \varphi_2)\\
  \end{array}
\]
Apart from the axioms, we also need \emph{modus ponens} and necessitation rules.

@@

%% -----------------------------------------------------------------------------

It's worth to mention about the choice of Hilbert-style systems as the deductive system in our library. The main reason is itself the soundness and completeness proofs: deductive systems such as natural deduction~\cite{Russo_PMLD, basin.ea:labelled:1997, SVRS02-recipe}, sequent calculi~\cite{Governatori00labelledmodal, 03-MSSV-fiblog19}, or tableaux~\cite{BrodaRusso05}, in modal logics, usually need the aid of labels or provisos. This makes, for instance, the metatheorem of soundness harder to proof: while Hilbert systems need only a simple structural induction on the size of the proof, labeled systems need to deal with several corner cases. The disadvantage of the Hilbert system is to use it as a way to establish proofs \emph{in} the system itself, which is not ``direct'' or ``natural''. However, it is still possible to use the $\mathcal{L}$tac language in Coq to automatize parts of this work.

%% -----------------------------------------------------------------------------

Finally, a central aspect in the study of modal logics is the so-called \emph{frame
  correspondence}, which is a relationship between the
accessibility relation and the truth of certain formulas in frames containing
it. As an example, consider a frame $\mathcal{F} = \langle W,R
\rangle$ in which the relation $R$ is reflexive: $\forall w \in W. (wRw)$, i.e.,
every world is accessible from itself. In such frame when $\Box A$ holds in
a world $w$, $w$ itself is among the accessible worlds in which $A$ must be true.
In this way, if the accessibility relation $R$ is reflexive, then whatever world
$w$ and formula $\varphi$ we take, the formula $\Box \varphi \to \varphi$ will
be true there. Examples of properties that characterize certain formula schemas are
presented in the table below.

@@

In this section, we describe the implementation details of our modal logic
library in Coq. We start by describing the structure of formula syntax and the
representation of frames and models, needed to define the Kripke semantics.
Next, we define a modular Hilbert-style deductive system. The modularization is
obtained by parameterizing it by a set of axioms used. This allows us to
generalize results about derivations to several deductive systems, independently
of the axioms used. We conclude by discussing the soundness, completeness, and the frame
correspondence proofs.

@@

To implement our library, we use a \emph{deep embedding} of
formulas, i.e., we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive formula: Set :=
| Lit    : nat -> formula
| Neg    : formula -> formula
| Box    : formula -> formula
| Dia    : formula -> formula
| And    : formula -> formula -> formula
| Or     : formula -> formula -> formula
| Implies: formula -> formula -> formula.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logic. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$. Furthermore, the type \mintinline{coq}{theory} is defined as a function
between \mintinline{coq}{formula} and Coq's \mintinline{coq}{Prop}s, which is used to represent a set of formulas.

@@

Before defining the Kripke semantics, we need to represent the model used to
interpret a modal logic formula. We encode a frame by a record formed by
a type \mintinline{coq}{W: Type} to represent the set of worlds and an
accessibility relation denoted by a type \mintinline{coq}{R: W -> W -> Prop}.
\begin{minted}{coq}
Record Frame: Type := {
  W: Type;
  R: W -> W -> Prop
}.
\end{minted}
A model is then denoted by another record, formed by a frame and a labeling function,
which assigns the truth value of a variable at a given world.
\begin{minted}{coq}
Record Model: Type := {
  F: Frame;
  v: nat -> (W F) -> Prop
}.
\end{minted}
Note that the type of the labeling function \mintinline{coq}{v} ensures that it
refers to a world from the set of possible worlds in the frame
\mintinline{coq}{F}\footnote{We recall that \mintinline{coq}{W F} is Coq's syntax for
accessing field \mintinline{coq}{W} from the record \mintinline{coq}{F}.}. The encoding of the Kripke semantics is done by a Coq function
that implements the relation of Figure~\ref{fig:kripke}; we omit the code
for propositional logics connectives for brevity. The interesting parts are for
variables (constructor \mintinline{coq}{Lit}) which uses the labeling function
\mintinline{coq}{v} from the model \mintinline{coq}{M}. The meaning of
modalities $\Box$ and $\Diamond$ is represented by quantifying over worlds
using the accessibility relation $R$.
\begin{minted}{coq}
Fixpoint formula_valuation (M: Model) (w: W (F M)) (phi: formula): Prop :=
  match phi with
  | Lit x
     => v M x w (* should be inhabited, i.e., true *)
  | Box psi
     => forall w': W (F M), R (F M) w w' -> formula_valuation M w' psi
  | Dia psi
     => exists2 w': W (F M), R (F M) w w' & formula_valuation M w' psi
  (* some code omitted *)
  end.
\end{minted}
In order to ease the task of writing formulas and its semantics properties, we
define Coq notations that mimic the standard way of writing them
by using Coq's custom notation features.

@@

A key property of the Hilbert deductive system as formalized
in our library is its soundness (through system K). If a formula $\varphi$ is derived from a
context $\Gamma$, i.e., $\Gamma\vdash\varphi$, then $\varphi$ is entailed within the deep embedding
from the set of hypotheses $\Gamma$, i.e., $\Gamma\models\varphi$. The Coq statement for the soundness
result is given as follow:

\begin{minted}{coq}
Theorem soundness: forall G p, (K; G |-- p) -> (G ||= p).
\end{minted}
